Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-17 paper/main.tex
fix unicode chat U+2212 error
mengyan reviewed sec 4.2
mengyan edit on several comments

\newcommand{\mengyan}[1]{\textcolor{magenta}{#1}}
\newcommand{\maciej}[1]{\textcolor{blue}{#1}}

@@

\bibliography{ref.bib}
\DeclareUnicodeCharacter{2212}{-}
\begin{document}

@@

Machine learning has been also used for prediction in proteins \cite{Yang2018}.
However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
The challenges of applying machine learning techniques on synthetic biology have two folds.
On the one hand, there is no large-scale and high-quality labelled data. The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
One the other hand, it is time-consuming and expensive to measure the TIR for RBS sequences \mengyan{Maciej: can you add more details or reference here}. 
We addressed the challenges by applying DBTL workflow. 
Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
\mengyan{why? we need to be more precise. Mengyan can add it later. Make it a paragraph, showing the challenges in learning with small number of noisy data and exploration-exploitation balance, and explain how we address that.}

We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2012parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
The results were then fed back to the algorithm for it to recommend the next round of designs.
This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.

The \textbf{Story} (3 sentences):\\
\begin{enumerate}
    \item we automate the design of sequences for finding high TIR RBS
    \begin{enumerate}
        \item DBTL
    \end{enumerate}
    \item the ML method discovers biological knowledge (validates existing knowledge), including the highest TIR ever
    \begin{enumerate}
        \item high G
        \item diverse library
        \item edit distance > 2
    \end{enumerate}
    \item BioFoundry helps ML (good data), and ML helps BioFoundry (more automation)
    \begin{enumerate}
        \item Sequences have small technical variation
        \item UCB much more efficient than random
    \end{enumerate}
\end{enumerate}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.4]{plots/Main_Paper/swarmplot.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/quantplot.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{\textbf{TIRs of RBS groups examined in this study.} 
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.
    \mengyan{Maciej: make the sentence more clear?}
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively. 
    SD - Shine-Dalgarno sequence. Dash line is set 1 and represents the averaged benchmark sequence TIR for that group. 
    \mengyan{explain why reference sequences are not all one.}
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group. 
    \mengyan{Maciej: maybe need to address why there are different groups in different plot, e.g. why no BPS-NC.}
    \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). The size of the circle represents the experimentally obtained TIR for measured groups.}
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
\mengyan{explain the TIR ratio we used.}
The Bandit-0 group performed poorly, due to being trained based on literature data.
\mengyan{Maciej: can you add some analysis on literature data?
to verify 1) our library is "better" 2) it's hard to do ML on available library. some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.

@@

Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
% Interestingly there wasn't much difference in fitness parameters between Round 1-3.
One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
\mengyan{Mengyan to explain more about the prediction v.s. recommendation.}

\mengyan{Some other points need to address: 
1) the TIR label we used here is different from the TIR ratio in Fig 2. 
2) the setting of predictor.}

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF.pdf}
    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each of the rounds. R\textsuperscript{2}, Spearman Correlation coefficient and p value are all provided for each plot.}
    \label{fig: Scatterplot}
\end{figure}

@@

There is a number of issues that we would like to address in the future.
The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
Essentially, there are two possible reasons for this, one is due to the algorithm and one due to nature.
If the problem is with the algorithm, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
It might be also possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
In that case, changes in the algorithm will not help.
Another problem we would like to tackle is the not always good prediction of the TIR of a given RBS.
\mengyan{Mengyan to address exploration purpose here.}
In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.

>>>>

\newcommand{\cheng}[1]{\textcolor{purple}{{\bf Cheng:~}#1}}
\newcommand{\mengyan}[1]{\textcolor{magenta}{#1}}
\newcommand{\maciej}[1]{\textcolor{blue}{#1}}

@@

\bibliography{ref.bib}
% \DeclareUnicodeCharacter{2212}{-}
\begin{document}

@@

% Machine learning has been also used for prediction in proteins \cite{Yang2018}.
However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
The challenges of applying machine learning techniques on synthetic biology have two folds.
On the one hand, there is no large-scale and high-quality labelled data. 
The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
And it is time-consuming and expensive to measure the TIR for RBS sequences. \mengyan{Maciej: can you add more details or reference here}.
One the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.  
We addressed the challenges by applying DBTL workflow. 
Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. 
% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\

We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2012parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
The results were then fed back to the algorithm for it to recommend the next round of designs.
This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.
\cheng{I suggest reporting everything that is 0.95 and above as "high" (assuming the ratio normalized TIR scale)}

The \textbf{Story} (3 sentences):
\begin{enumerate}
    \item we automate the design of sequences for finding high TIR RBS
    \begin{enumerate}
        \item DBTL
    \end{enumerate}
    \item the ML method discovers biological knowledge (validates existing knowledge), including the highest TIR ever
    \begin{enumerate}
        \item high G
        \item diverse library
        \item edit distance $>$ 2
    \end{enumerate}
    \item BioFoundry helps ML (good data), and ML helps BioFoundry (more automation)
    \begin{enumerate}
        \item Sequences have small technical variation
        \item UCB much more efficient than random
    \end{enumerate}
\end{enumerate}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.4]{plots/Main_Paper/swarmplot.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/quantplot.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{\cheng{May need to use subfigure to label the figures as A,B,C,D. I recall a discussion that C colours with viridis makes it hard to see the different histograms. But let's keep to the viridis colours. Also, if it is not too hard, make the bin edges [0, 0.1, 0.2, ...].}
    \textbf{TIRs of RBS groups examined in this study.} 
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.
    \mengyan{Maciej: make the sentence more clear?}
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively. 
    SD - Shine-Dalgarno sequence. Dash line is set 1 and represents the averaged benchmark sequence TIR for that group. 
    \mengyan{explain why reference sequences are not all one.}
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group. 
    \mengyan{Maciej: maybe need to address why there are different groups in different plot, e.g. why no BPS-NC.}
    \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). The size \cheng{(area or diameter?, should be area)} of the circle represents the experimentally obtained TIR for measured groups.}
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
We measure the TIR of reference RBS as experimental control. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of reference sequences in each round.
The Bandit-0 group performed poorly, due to being trained based on literature data.
\mengyan{Maciej: can you add some analysis on literature data?
to verify 1) our library is "better" 2) it's hard to do ML on available library. some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.

@@

Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
Note the TIR values reported in Figure \ref{fig: Scatterplot} is normalised TIR based on the standardisation descried in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
% Interestingly there wasn't much difference in fitness parameters between Round 1-3.
% One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
One interesting point is that the predictions are also influenced by our recommendation choices. 
In each round, we selected parts of points for exploration (with high predicted uncertainty), which means the predictors are unlikely to give accurate predicted TIR mean for those points due to lack of observations in data points lying in those areas. 
However, for those explorative points, although the predictors give poor prediction for the predicted TIR, the recommendation choices do allow the predictor to gain new information and learn about those unexplored areas.

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF.pdf}
    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm (GPR with WDS kernel, with maximum kmer as 6 and maximum shift as 1) calculated after each of the rounds.
    The parameters settings are 
    R\textsuperscript{2}, Spearman Correlation coefficient and p value are all provided for each plot.}
    \label{fig: Scatterplot}
\end{figure}

@@

There is a number of issues that we would like to address in the future.
The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
Essentially, there are two possible reasons for this, one due to nature and one is due to the algorithm.
One the one hand, it might be possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
In that case, improves in the algorithm will not help.
On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
% In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
% Another problem we would like to tackle is the not prediction of the TIR of a given RBS.
% In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
There are open questions need to be addressed for applying machine learning in synthetic biology in general.
For example, 
how can we understand the language of RBS sequences better?
Given a small amount of RBS sequences, how can machine learning algorithm provides more accurate predictions and uncertainty measurement? 
How should we choose the parameters which controls the exploration-exploitation rate?

### 2021-05-17 paper/methods.tex
mengyan edit on several comments

\subsubsection{Data pre-processing}

@@

The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
\mengyan{maybe we need to introduce spectrum kernel as well since we used it for round 0.}
To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.  
WDS is a type of string kernels, which take two sequences as inputs and output a scalar value which represents the similarities between the two sequences.  
WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
The maximum substring length is specified by $\ell$. 
The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
This could be useful when there is a shift between two sequences.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. 
% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.

>>>>

\subsubsection{Data pre-processing}
\label{sec: method data pre-procesing}

@@

The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.  
WDS is a type of string kernels, which take two sequences as inputs and output a scalar value which represents the similarities between the two sequences.  
WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
The maximum substring length is specified by $\ell$. 
The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
This could be useful when there is a shift between two sequences.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. 
% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.

### 2021-05-17 paper/supplementary.tex
fix bugs for local env
add challenges in result sec
update paper plots
Updates from Overleaf
mengyan reviewed sec 4.2
mengyan edit on several comments

\begin{figure}
    \centering
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
    \caption{Kernel Heatmap and Predictions. Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265). Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right). The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
    \label{fig:my_label}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
    \caption{TSNE of RBS sequences in design space with clustering. The distance is calculated based on the weighted degree kernel on RBS sequences. Colours indicate different groups, shapes indicate different clusters.}
    \label{fig:my_label}
\end{figure}

\subsection{Gaussian Process Regression}

@@

\subsubsection{Normalisation of Kernel}

@@

For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
%which is based on the \textit{optimism in the face of uncertainty}, 
selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
\begin{align}
\label{Eq: GPUCB}
    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
\end{align}
where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.

@@

\subsection{Design Pipeline}

The n + 1 round design is based on the nth round result, where each sequence has 6 replicates with TIR labels.
We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth. 
After normalisation, each replicate has zero mean and unit variance.

For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
We model the covariance matrix using the weighted degree kernel with shift. 
We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation. 
We choose $l=6, s= 1, \alpha = 2$ for the second round design.

For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
We recommend 90 sequences from the design space.

\subsection{Intuition behind UCB and visualisation}

\begin{itemize}
    \item Exploitation and exploration explanation.
    \item Visualise coverage by clustering plot.
    \item Table for in-clustering mean and variance.
\end{itemize}

\subsection{Result analysis}

\begin{itemize}
    \item violinplot.
    \item regression performance plot, table.
    \item kernel matrix plot.
\end{itemize}

@@

\input{paper/mixture_absGau}

>>>>

\begin{figure}
    \centering
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
    \caption{Kernel Heatmap and Predictions. Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265). Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right). The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
    \caption{TSNE of RBS sequences in design space with clustering. The distance is calculated based on the weighted degree kernel on RBS sequences. Colours indicate different groups, shapes indicate different clusters.}
\end{figure}

\subsection{Gaussian Process Regression}
\label{supp: GPR}

@@

\subsubsection{Normalisation of Kernel}
\label{supp: Normalisation of Kernel}

@@

For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
%which is based on the \textit{optimism in the face of uncertainty}, 
selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
\begin{align}
    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
    \tag{\ref{Eq: GPUCB}}
\end{align}
where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.

@@

\subsection{Machine Learning Design Pipeline}
In this section, we report in details about our machine learning design pipeline.
The raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.

\mengyan{explain somewhere why we have a different pipeline for round 0.}\\
\textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.
We first normalise the raw TIR to values between 0 and 1. 
We applied the Gaussian Process Regression with noise parameter $\alpha = 1e-10$. 
We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
The design size is 60 with UCB parameter $\beta = 1$.\\
\textbf{Bandit-1 to Bandit-3: }
the t + 1 round design is based on the $t^th$ ($t \geq 1$) round result, where each sequence has 6 replicates with TIR labels.
We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively.
After normalisation, each replicate has zero mean and unit variance. 
For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
We model the covariance matrix using the weighted degree kernel with shift. 
We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation. 
We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
We recommend 90 sequences from the design space.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{plots/Supplementary/All_logo.pdf}
    \caption{\textbf{Sequence logo for all tested sequences.} }
    \label{fig: All_logo}
\end{figure}

\input{mixture_absGau}
